---
title: "Assignment 5: Data Visualization"
author: "Tasneem Ahsanullah"
date: "Spring 2023"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1.  Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Upload the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 
#importing tidyverse, lubridate, here & cowplot 
library(tidyverse); library(lubridate); library(here); 
library(cowplot)

#Assigned a variable to the processed data folder location
processed_data = "Data/Processed"

#reading in the data
NTL.chem.PeterPaul <- read.csv(
  here(processed_data,
  "NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv"),
  stringsAsFactors = TRUE)

NTL.chem.phys.PeterPaul <- read.csv(
  here(processed_data,
  "NTL-LTER_Lake_ChemistryPhysics_PeterPaul_Processed.csv"),
  stringsAsFactors = TRUE)

NTL.nutrients.PeterPaul <- read.csv(
  here(processed_data,
  "NTL-LTER_Lake_Nutrients_PeterPaul_Processed.csv"),
  stringsAsFactors = TRUE)

NEON.litter <- read.csv(
  here(processed_data,
  "NEON_NIWO_Litter_mass_trap_Processed.csv"),
  stringsAsFactors = TRUE)

#2 
#checking if dates are in date format
class(NTL.chem.PeterPaul$sampledate)
class(NTL.chem.phys.PeterPaul$sampledate)
class(NTL.nutrients.PeterPaul$sampledate)
class(NEON.litter$collectDate)

#changing format to date
NTL.chem.PeterPaul$sampledate <- ymd(NTL.chem.PeterPaul$sampledate)

NTL.chem.phys.PeterPaul$sampledate <- ymd(NTL.chem.phys.PeterPaul$sampledate)

NTL.nutrients.PeterPaul$sampledate <- ymd(NTL.nutrients.PeterPaul$sampledate)

NEON.litter$collectDate <- ymd(NEON.litter$collectDate)

```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#custom theme with adjusted plot background, title and legend background/title
lake.theme <- theme_classic() + 
  theme(
    plot.background = element_rect(
      color='black',
      fill='seashell'
    ),    
    plot.title = element_text(
      colour = 'black',
      size = '14'
    ),
    legend.background = element_rect(
      color='black',
      fill ='white'
    ),
    legend.title = element_text(
      color='black'
    )
    
  )

#setting the theme for all plots
theme_set(lake.theme)
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 
#phosphorous by phosphate plot differentiated by lake
#with a line of best fit
phos.plot <- NTL.chem.PeterPaul %>% 
  ggplot(aes(
      x=po4,
      y=tp_ug,
      color=lakename)
    ) + 
  geom_point(size=0.5,alpha=0.5) +
    labs(
    title="Phosphorous/Phosphate for Peter & Paul Lakes",
    x="Phosphate (po4)",
    y="Phosphorous (ug)",
    color="Lakename"
    ) +
  geom_smooth(
    method='lm',
    se=FALSE,
    color='Black') +
  xlim(0,50)

print(phos.plot)
```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tip: R has a build in variable called `month.abb` that returns a list of months;see <https://r-lang.com/month-abb-in-r-with-example>

```{r Create boxplots}
#5 
#boxplot for temperature by month with lake name as the color aesthetic
temp.plot <-
  ggplot(NTL.chem.PeterPaul, 
    aes(x=factor(month,levels=1:12,labels=month.abb), 
    y = temperature_C)) +
  geom_boxplot(aes(color = lakename)) +
labs(
  title="Temperature for Peter & Paul Lakes",
    x="Month",
    y="Temperature (C)",
    color="Lakename"
  )
print(temp.plot)

#boxplot for phosphorous by month with lake name as the color aesthetic
tp.plot <-
  ggplot(NTL.chem.PeterPaul, 
  aes(x=factor(month,levels=1:12,labels=month.abb),
  y = tp_ug)) +
geom_boxplot(aes(color = lakename)) +
labs(
  title="Phosphorous for Peter & Paul Lakes",
    x="Month",
    y="Phosphorous (ug)",
    color="Lakename"
  )
print(tp.plot)

#boxplot for nitrogen by month with lake name as the color aesthetic
tn.plot <-
  ggplot(NTL.chem.PeterPaul, 
  aes(x=factor(month,levels=1:12,labels=month.abb),
  y = tn_ug)) +
geom_boxplot(aes(color = lakename)) +
labs(
  title="Nitrogen for Peter & Paul Lakes",
    x="Month",
    y="Nitrogen (ug)",
    color="Lakename"
  )
print(tn.plot)

plot_grid(temp.plot, tp.plot, tn.plot, ncol= 1, 
          align='v',axis='b')



```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: The variables of interest increase in the spring, peak in summer and then decrease in winter. 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6
#plot of the Needle dry mass by date
litter.plot <- NEON.litter %>% 
  filter(functionalGroup=='Needles') %>% 
  ggplot(aes(
      x=collectDate,
      y=dryMass,
      color=nlcdClass)
    ) + 
  geom_point(size=0.5,alpha=0.5) +
    labs(
    title="Dry Mass of Needle Litter",
    x="Date",
    y="Dry Mass",
    color="NLCD Class"
    )

print(litter.plot)

#7
#plot of the Needle dry mass by date for each NLCD class
litter.facet <- NEON.litter %>% 
  filter(functionalGroup=='Needles') %>%
  ggplot(aes(
      x=collectDate,
      y=dryMass,
      )
    ) + 
  geom_point(size=0.5,alpha=0.5) + 
    facet_wrap(vars(nlcdClass)) +
    labs(
    title="Dry Mass of Needle Litter by NLCD",
    x="Date",
    y="Dry Mass"
    )
print(litter.facet)

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: Plot 6 is more effective because it is easier to compare the different NLCD to each other since they are on the same graph and just different colors. 