---
title: "Assignment 2: Coding Basics"
author: "Tasneem Ahsanullah"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics Day 1

1.  Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r}
#1. 
seq4 <- seq(1,100,4) #assigning a sequence of 1-100 increasing by 4s to the variable seq 4
seq4 #displaying the sequence of 1-100 by 4s

#2. 
mean(seq4) #computing the mean of the sequence of 4s. (49)
median(seq4) #computng the median of the sequence of 4s. (49)

#3. 
mean(seq4)>median(seq4) #inequality testing whether the mean of the sequence of 4s is greater than the median. Answer is False so they are not equal
```

## Basics Day 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r}
name <- c('carl', 'sam', 'emma', 'chloe', 'ella') #character vector
scores <- c(67,44,89,38,73) #numeric vector
passed <- c('TRUE','FALSE','TRUE','FALSE','TRUE') #logical vector
class.grades.df <- data.frame("Student Name"=name, "Test Score" = scores, "Passed (50 or higher)"=passed)
class.grades.df
```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: this data frame is different from a matrix because it has more than one class of data.

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r}
Passing <- ifelse(scores>50,"TRUE","FALSE")
print(Passing)
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: ifelse worked instead of if and else because scores is a numeric vector and you would need a logical vector for if and else.
