---
title: 'Assignment 4: Data Wrangling'
author: "Tasneem Ahsanullah"
date: "Spring 2023"
output:
  pdf_document: default
  word_document: default
geometry: margin=2.54cm
editor_options:
  chunk_output_type: console
---



## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Wrangling 

## Directions

1.  Rename this file `<FirstLast>_A04_DataWrangling.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

The completed exercise is due on Friday, Feb 20th @ 5:00pm.

## Set up your session 

1a. Load the `tidyverse`,  `lubridate`, and `here` packages into your session.

1b. Check your working directory.

1c. Read in all four raw data files associated with the EPA Air dataset, being sure to set string columns to be read in a factors. See the README file for the EPA air datasets for more information (especially if you have not worked with air quality data previously). 

2. Apply the `glimpse()` function to reveal the dimensions, column names, and structure of each dataset. 

```{r, message = FALSE}
#1a 
#loading tidyverse, lubridate and here packages to the session
library(tidyverse)
library(lubridate)
library(here)

#1b 
#retrieving the working directory
getwd()

#1c 
#reading in all four EPA Air datasets, assigning a name and setting string columns to
#be read as factors
Air_O3_2018 <- read.csv("./Data/Raw/EPAair_O3_NC2018_raw.csv",
  stringsAsFactors = TRUE)

Air_O3_2019 <- read.csv("./Data/Raw/EPAair_O3_NC2019_raw.csv",
  stringsAsFactors = TRUE)

Air_PM25_2018 <- read.csv("./Data/Raw/EPAair_PM25_NC2018_raw.csv",
  stringsAsFactors = TRUE)

Air_PM25_2019 <- read.csv("./Data/Raw/EPAair_PM25_NC2019_raw.csv",
  stringsAsFactors = TRUE)

#2 
#using glimpse to get the dimensions, column names, and structure of each dataset 
glimpse(Air_O3_2018)
glimpse(Air_O3_2019)
glimpse(Air_PM25_2018)
glimpse(Air_PM25_2019)

```


## Wrangle individual datasets to create processed files.

3. Change date columns to be date objects.

4. Select the following columns: Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE

5. For the PM2.5 datasets, fill all cells in AQS_PARAMETER_DESC with "PM2.5" (all cells in this column should be identical).

6. Save all four processed datasets in the Processed folder. Use the same file names as the raw files but replace "raw" with "processed".

```{r wrangling}
#3
#setting date columns as objects
Air_O3_2018$Date <- mdy(Air_O3_2018$Date)
Air_O3_2019$Date <- mdy(Air_O3_2019$Date)
Air_PM25_2018$Date <- mdy(Air_PM25_2018$Date)
Air_PM25_2019$Date <- mdy(Air_PM25_2019$Date)

#4
#selecting Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, 
#SITE_LATITUDE, SITE_LONGITUDE in all four data sets
Air_O3_2018_select <- select(Air_O3_2018, Date, DAILY_AQI_VALUE, Site.Name,
    AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE)

Air_O3_2019_select <- select(Air_O3_2019, Date, DAILY_AQI_VALUE, Site.Name, 
    AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE)

Air_PM25_2018_select <- select(Air_PM25_2018, Date, DAILY_AQI_VALUE, Site.Name, 
    AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE)

Air_PM25_2019_select <- select(Air_PM25_2019, Date, DAILY_AQI_VALUE, Site.Name, 
    AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE)

#5
#filling all cells in AQS_PARAMETER_DESC with "PM2.5" for both PM25 datasets
Air_PM25_2018_select$AQS_PARAMETER_DESC <- "PM2.5"
Air_PM25_2019_select$AQS_PARAMETER_DESC <- "PM2.5"

#6
#saving the new files into the Processed folder using the write.csv command
write.csv(Air_O3_2018_select, file = 
    "./Data/Processed/EPAair_O3_NC2018_processed.csv", row.names=FALSE)

write.csv(Air_O3_2019_select , file = 
    "./Data/Processed/EPAair_O3_NC2019_processed.csv", row.names=FALSE)

write.csv(Air_PM25_2018_select, file = 
    "./Data/Processed/EPAair_PM25_NC2018_processed.csv", row.names=FALSE)

write.csv(Air_PM25_2019_select, file = 
    "./Data/Processed/EPAair_PM25_NC2019_processed.csv", row.names=FALSE)
```


## Combine datasets

7. Combine the four datasets with `rbind`. Make sure your column names are identical prior to running this code.

8. Wrangle your new dataset with a pipe function (%>%) so that it fills the following conditions: 

  * Include all sites that the four data frames have in common: "Linville Falls", "Durham Armory", "Leggett", "Hattie Avenue", "Clemmons Middle", "Mendenhall School", "Frying Pan Mountain", "West Johnston Co.", "Garinger High School", "Castle Hayne", "Pitt Agri. Center", "Bryson City", "Millbrook School" (the function `intersect` can figure out common factor levels - but it will include sites with missing site information...) 

  * Some sites have multiple measurements per day. Use the split-apply-combine strategy to generate daily means: group by date, site name, AQS parameter, and county. Take the mean of the AQI value, latitude, and longitude.
  
  * Add columns for "Month" and "Year" by parsing your "Date" column (hint: `lubridate` package)
  
  * Hint: the dimensions of this dataset should be 14,752 x 9.

9. Spread your datasets such that AQI values for ozone and PM2.5 are in separate columns. Each location on a specific date should now occupy only one row.

10. Call up the dimensions of your new tidy dataset.

11. Save your processed dataset with the following file name: "EPAair_O3_PM25_NC1819_Processed.csv"

```{r combining}
#7 
#combining the 4 processed data sets into one data set

combined_air <- rbind(Air_O3_2018_select,Air_O3_2019_select,Air_PM25_2018_select,Air_O3_2019_select)

#8 
#filtered for the common sites amongst all 4 data sets, grouped by date, site name,
#AQS parameter, and county, and used summarise/mean function to take the mean of AQI,
#latitude and longitude.
wrangled_air_data <- 
  combined_air %>%
  filter(Site.Name == "Linville Falls" | Site.Name == "Durham Armory" | 
        Site.Name == "Leggett" | Site.Name == "Hattie Avenue" | 
        Site.Name == "Clemmons Middle" | Site.Name == "Mendenhall School" | 
        Site.Name == "Frying Pan Mountain" | Site.Name == "West Johnston Co." |
        Site.Name == "Garinger High School" | Site.Name == "Castle Hayne" | 
        Site.Name == "Pitt Agri. Center" | Site.Name == "Bryson City" | 
        Site.Name == "Millbrook School") %>%
group_by(Date, Site.Name, AQS_PARAMETER_DESC, COUNTY) %>%
filter(!is.na(DAILY_AQI_VALUE) & !is.na(SITE_LATITUDE) & !is.na(SITE_LONGITUDE)) %>%
  summarise(meanAQI = mean(DAILY_AQI_VALUE), 
            meanLatitude = mean(SITE_LATITUDE),
            meanLongitude = mean(SITE_LONGITUDE))

#added a month and year column using the month and year lubridate function
wrangled_air_data <- mutate(wrangled_air_data, month = month(Date)) 
wrangled_air_data <- mutate(wrangled_air_data, year = year(Date)) 



#9
#spread datasets so AQI values for ozone and PM2.5 are in separate columns using #pivot_wider
spread.data<-
  wrangled_air_data %>% 
  pivot_wider(names_from = AQS_PARAMETER_DESC,
              values_from = meanAQI) 
#10
#dimensions of spread data
dim(spread.data)

#11
#saving processed dataset in processed folder using write.csv
write.csv(wrangled_air_data, row.names = FALSE, 
          file ="./Data/Processed/EPAair_O3_PM25_NC1819_Processed.csv")

```


## Generate summary tables

12. Use the split-apply-combine strategy to generate a summary data frame. Data should be grouped by site, month, and year. Generate the mean AQI values for ozone and PM2.5 for each group. Then, add a pipe to remove instances where mean **ozone** values are not available (use the function `drop_na` in your pipe). It's ok to have missing mean PM2.5 values in this result. 

13. Call up the dimensions of the summary dataset.

```{r}

#12
#mean AQI values for ozone and PM2.5 grouped by site, month, and year. Removed missing mean values for Ozone with drop_na.
summary_air_data <- 
  spread.data %>% 
  group_by(Site.Name, month, year) %>% 
  drop_na(Ozone) %>% 
  summarize(mean.ozone <- mean(Ozone),
            mean.PM25 <- mean(PM2.5))

#13
#dimensions of summary data
dim(summary_air_data)

```

14. Why did we use the function `drop_na` rather than `na.omit`?

> Answer: drop_na gets rid of the whole row if there is a missing value while na omit just gets rid of the single NA value
